---
title: "InSTRuCt project: simulated e-trigger data"
author: "Andrew Zimolzak"
output: pdf_document
---

The e-triggers for this project answer the question, "Which patients had a test that shows a
possibility of cancer but have *not* had timely follow-up?" Abnormal tests can be positive fecal immunochemical test,
or lung imaging suspicious for malignancy).

Each site runs the e-trigger monthly. As output, they see the number of
abnormal tests altogether (denominator), and the number
of abnormal tests *without* expected follow up (numerator).
From this, they can calculate and graph their rate of missed test results, to determine whether
their quality improvement activities are effective.
On each graph, lower values are
better (indicating fewer misses and more tests with timely follow-up).



# Simulating e-trigger data over time

```{r libraries, message=FALSE, warning=FALSE, include=FALSE}
library(ggplot2)
library(tidyr)
library(dplyr)
library(knitr)
library(here)
```



```{r functions, include=FALSE}
x_to_proportion <- function(x) { 
  0.5 + atan(x) / pi
  # max slope = 1/pi = 0.32, at x=0.
  # going up by 1 makes proportion go about 0.5 to 0.75
  # x up by 0.03 makes p go to 0.51
}

proportion_to_x <- function(proportion) { 
  tan(pi * (proportion - 0.5))
}

next_proportion <- function(current_proportion) {  # returns simple numeric
  current_x <- proportion_to_x(current_proportion)
  new_x <- current_x + rnorm(1) - 0.1  # decrease a little each month
  x_to_proportion(new_x)
}

extend_by_one <- function(proportion_vec) {  # returns numeric vector
  val_to_add <- next_proportion(tail(proportion_vec, n=1))
  c(proportion_vec, val_to_add)
}

generate_run <- function(n) {  # returns numeric vector of length n
  my_run <- c(next_proportion(0.6))  # starting value
  for (i in 2:n) {
    my_run <- extend_by_one(my_run)
  }
  my_run
}

generate_etrigger_data <- function(number_sites, months_per_site) {  # returns data frame
  result <- generate_run(months_per_site)
  site_labels <- rep(LETTERS[1], months_per_site)
  timepoints <- 1:months_per_site
  for (i in 2:number_sites) {
    result <- c(result, generate_run(months_per_site))
    site_labels <- c(site_labels, rep(LETTERS[i], months_per_site))
    timepoints <- c(timepoints, 1:months_per_site)
  }
  data.frame(Proportion_raw = result, Site = site_labels, Month = timepoints)
}
```

Set up some global parameters

```{r generate-dataframe, echo=TRUE}
n_sites <- 11
last_prework_month <- 6
n_action <- 6
n_sustain <- 6
last_action_month <- last_prework_month + n_action
months_per_site <- last_action_month + n_sustain
project_kickoff <- as.Date('2021-01-01')

set.seed(123423)

X_continuous <- generate_etrigger_data(n_sites, months_per_site)

```

We generate:

- A few months of baseline data, plus prework phase data (total `r last_prework_month` months)
- `r n_action` months action phase
- `r n_sustain` months sustain phase (in reality, it varies 0-12 months)

Therefore, we will use a total of *`r months_per_site` months* per site, times `r n_sites` sites.

For realism, data are simulated as a variety of "random walk." Specifically,
each month's rate depends on last month's rate, plus a random step of
reasonable size either up or down, with constraints to ensure that proportions
can approach but never reach 1 or 0. The denominator (number of abnormals per month, with
or without follow-up) is modeled from a Poisson distribution, where each site has
a typical number of abnormals per month, with natural variation around the mean each month,
but no variation in site-specific mean over time.

```{r rational-numbers, include=FALSE}
expected_n = data.frame(
  Site = LETTERS[1: n_sites],
  expected = runif(n_sites, 9, 30)  # Sites usually 9 to 30 abnormal tests per mo.
)

rpois_vec <- Vectorize(rpois, 'lambda')

inner_join(X_continuous, expected_n, by='Site') %>%
  mutate(
    Denominator = rpois_vec(1, expected),
    Numerator = round(Denominator * Proportion_raw),
    Proportion = Numerator / Denominator,
    Phase = case_when(
      Month <= 3 ~ 'baseline',
      Month <= last_prework_month ~ 'prework',
      Month <= last_action_month ~ 'action',
      TRUE ~ 'sustain'
    ),
    Cohort = case_when(
      Site %in% c('A', 'B', 'C', 'D') ~ 1,
      Site %in% c('E', 'F', 'G', 'H') ~ 2,
      TRUE ~ 3,
    ),
    Date = n_action * 30 * (Cohort - 1) + project_kickoff
  ) %>%
  select(Cohort, Site, Month, Phase, Numerator, Denominator, Proportion, Date) -> X
```

We will also compute the average rate of misses by month, across *all* sites.

```{r compute-average, include=FALSE}
X %>%
  group_by(Month) %>%
  summarise(Proportion = mean(Proportion)) %>%
  mutate(Site='p_Avg') ->
  averaged
site_data_with_average <- bind_rows(X, averaged)
```




# Results

```{r plotting, echo=FALSE}
  ggplot(site_data_with_average, aes(Month, Proportion)) +
    geom_line() +
    geom_vline(xintercept = last_prework_month) +
    geom_vline(xintercept = last_action_month) +
    ylim(0,1) +
    facet_wrap(vars(Site))
```

In the figure above, the vertical reference lines mark the start and end of the
action phase (period when the site was interacting with the central study team
to implement quality interventions). Before the first line is the baseline and
"prework" phase. After the second line is the "sustainment" phase.

Each panel is a different VA medical center site. The lower right panel shows
the average rate of misses for all 11 sites.




# Sample data table

```{r sample-data, echo=FALSE}
head(X, 18+5) %>% kable()

output_filename <- 'fake_data.csv'
write.csv(X, file=here(output_filename), row.names = FALSE)
Xrows = dim(X)[1]
Xcols = dim(X)[2]
```

Wrote output CSV (`r Xrows` rows $\times$ `r Xcols` columns) to the file
``r output_filename``, in the following directory:

``r here()``


# Plot design

```{r dots-no-facet}
site_data_with_average %>%
  filter(Site != 'p_Avg') %>%
  ggplot(aes(as.factor(Month), Proportion)) +
    geom_boxplot() +
    geom_jitter(width = 0.1, alpha = 0.5) +
    geom_vline(xintercept = last_prework_month) +
    geom_vline(xintercept = last_action_month) +
    ylim(0,1) +
    labs(y="Proportion of abnormal FOBT missed", x='Month (start date aligned to 6)')
```


